draft 1

### 2016-06-27 finalwriteup/draft.rmd
started TP FP

![This method was implemented by the British government for a rail project which is still under completion](UKrail.JPG)

@@

The following chapter, Chapter 4, describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting return per dollar, followed by a simpler binary classification problem predicitng profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision making scenarios and further applications were considered.

Chapter 5 presents the results of the series of analyses described above and directly relates them to the overall aim and hypotheses of the project. This is in the format of tables and plots where results are briefly discussed. The examination of results is taken deeper as well as in a broader context in Chapter 6: Discussion. Here the nature of the results are debated in terms of how surprising they were, their impact on the research hypothesis and limitations as well as suggested future work. The final Chapter 7: Conclusion summarises the  findings of the project, how the research answered the hypotheses and how the work could be viably applied in industry.

@@

```{r}
## Method Conclusion

# In summary, to determine whether a 
# * obtained data: 4169 jobs from a single case study company over 12 years. Data was de-dentified by an employee of the company
# * the data was all from the company CRM. There were three chief sources:
#     * invoicing data
#     * timesheet data
#     * project summary data
# * the data was cleaned from mistakes which were generally discovered as outliers 
# * many new variables were engineered from the data, often by summarising time series data into a value that is relevant to a final data set of a single job per row
# * variable selection: dozens of variables available which were condensed to the most important 10-12. variable importance methods were ANOVA, random forest, and cforests. the overlapping important variables were chosen first as well as some others
# * 5 models were tested: baseiline: log reg, naiive bayes.
# * sophisticated ML: boosted reg, random frest, bayesian networks
# * best were comapred to a model averaged version
# * Analysis of impact on bottom line for business if model is integrated into decision making
# * link: results
```

# Results

This chapter presents key charts and tables of results that progressively answer the research question: can a statistical or machine learning model based on a company's internal historical project data predict the profitability of a new project and have a positive impact on the bottom line of the business?

@@

@@

### Random Forests

@@

### Cforest

@@

Once several methods of variable importance analysis have been completed, the separate results were rationalised and final variables were selected for the predictive models.

### Variable Selection Summary

Based on the results of the three methods for variable importance, 11 independent variables were selected for predictive modelling. They are listed in the table below along with the methods that highlighted each variable as important (randomForest = RF, cForest = CF):

@@

The variables above were used to build the predictive models and the results are summarised in the following two sections.

## Model Results

Once the explanatory variables were narrowed down to a succinct list, the predictive models were built and tested. First, prediction of return per dollar as a regression problem was attempted followed by prediction of profit or loss as a binary classification problem. linear regression, random forests and boosted tree methods were applied to both problems and had varying levels of success.

### Regression Models

@@

were statistically significant. Two types of ANOVA models were run. The first used only a 'core' list of 6 variables that were largely complete and any relevant interaction terms. The second method was developed to overcome the problem of missing data. When all variables were combined, only 15% of the project cases had complete data and therefore the ANOVA test would eliminate 85% of the data upon execution. Imputing the missing data was not desirable at that stage due to the volume of missing categorical data and an inability to ensure the imputed data was not hampering results. The second method was a more complicated procedure that required dozens of models but allowed variables with missing data to be included. First, a project was randomly selected (a row in the dataset) which would have certain variables complete. The data was then filtered for all projects that had at least the same columns complete, and an ANOVA model was built on this reduced data set. This method allowed variables with a larger proportion of missing data to be included while not attempting to analyse all variables at once.

@@

results.A2 = readRDS("reg_lm.rds")
results.A2 <- dplyr::filter(results.A2, diff>-15)
# n.samples(data = results.A2$diff)

@@

```{r load_rf_results}
results.rf.core = readRDS("reg_rf_core.rds")
# n.samples(data = results.rf.core$diff)

results.rf2 = readRDS("reg_rf.rds")
# n.samples(data = results.rf2$diff)

@@

The two regression models predicting return per dollar resulted in high mean RMSE's ranging from \$`r mean(results.rf.core$final.RMSE) %>% round(2)` to \$`r mean(results.A2$final.RMSE) %>% round(2)`. Because of the poor results, further development of regression models was deffered in favour of investiation of binary classification models. Binary classification is a simpler task than regression because it requires less precision. The challenge is how to gainfully apply the results to the business scenario.

### Binary Classification Models

The cost estimation problem analysed in the literature review had been converted to a profitability estimation problem and was finally converted to a binary profit/loss prediction. The predictive formula was re-structured so that the 11 explanatory variables predicted a new response variable: profitable (or not). The original return per dollar response variable was transformed by converting aLl 'return per dollar' values less than or equal to zero to 'loss making' jobs and 'return per dollar' values greater than 1 became 'profitable' jobs. The following section summarises results from the trialled binary classification methods.

#### Initial Trial of Five Methods

The five prediction methods, Boosted Trees, Naive Bayes, Logistic Regression, Random Forests and Bayesian Networks, were intially trialled without imputing data, using the maximum amount of data possible depending on the method. Boosted Trees and Naiive Bayes are able to process data with missing values, so all data could be input [@gbm; @e1071]. Logistic Regression, Random Forest and Bayesian Networks require datasets without missing values so a subset of complete data was used for these. It was expected that Boosted Trees would perform best as they are known to have powerful predictive capabilities and can use incomplete data. For this reason, 4 x 5-fold runs of each method were performed (20 models each) and statistically compared against Boosted Trees' results. A power calculation was then used to calculate the number of runs required for a statistical power of 0.8.

@@

Before the plot could be made, the Boosted Tree algorithm parameters were tuned to optimise the AUC outcomes. The caret package was used for this and the tuned parameters are summarised below [@caret]:

@@

```{r}
# image below generated from file rf_log_naiive_auc.rmd
# lines 683 to 689

@@

reduc.totAUC <- readRDS("reduc_df_auc.rds")

@@

Results using imputed data tightened the distribution of randomForest AUC output substantially but slightly increased the distribution spread of Boosted Trees and Logistic Regression. Since the methods using imputed data performed similarly, substantially tightened the distribution of randomForest results, and made use of more data, the imputed data models were propogated forward to the development of blended models.

#### Blended Models

Blended models combine the predictions from each higher-performing individual model to create averaged or 'blended' predictions. In this case, the randomForest, Boosted Trees, and Logistic Regression models were chosen and as outlined in the method, 6 techniques for model blending were trialled ranging from simple to complex. The simplest way was to average the probability output (a number between 0 and 1) of all three models. The next two simplest methods were to build a logistic regression and boosted tree from the three model outputs only. Feature Weighted Linear Stacking (FWLS), randomForest and boosted trees were also tested as blending methods where the predictions from each model became additional variables to the original explanatory variables (meta-features). The model predictions were interacted with each original variable so that models that performed more strongly under certain meta-feature conditions could be weighted as such. Some meta-features may have been dispensable in the blending process, so in order to maintain simple comprehensible models, a variable importance study was conducted again.

##### Variable Selection: Blended Model

@@

The logistic regression found the F-statistics from the following variables were significant in rejecting the null hypothesis (p-value < 0.05):

@@

* Discipline * results of logistic regression model
* Discipline * results of boosted tree model
* team size for project * results of logistic regression model
* Client Business type * results of boosted tree model

@@

* Total amount invoiced for the job
* Timespan
* Project category

@@

#### Comparison of Blended Models

@@

auc.list <- readRDS("report_data/blend_auc1.rds")
means = lapply(auc.list, mean) %>% unlist
greater = means[means> means["orig.log"]] %>% names
auc.melt = melt(auc.list)
auc.melt$greater = ifelse(auc.melt$L1 %in% greater, TRUE, FALSE)

@@

```{r, fig.width = 8, fig.height=10}

@@

### Bottom Line Analysis

The first research question, whether machine learning and statistical techniques can predict project profitability based on historic data, has been answered. The measure of profitability could not be modelled but simplifying the profitability prediction to a binary (profit/loss) model was successful. To answer the second research question, whether the model would improve the business' profitability, a bottom line analysis was performed.

The models output a 'probability' between 0 and 1 that each project will be a loss-making job (where probability = 1). This leaves the question, at what probability would a decision-maker round the probability up to 1 or down to 0? And what business decision would then be made? To find the rounding threshold, an experimental business-scenario was tested. At a given threshold, all projects calculated above the threshold were considered too risky, and were rejected. All profits and losses from these projects were forfeited, and the profits from the remaining jobs (below the threshold) were summed to give a total profit.

@@

# How to achieve statistical power of 0.8?
finance.plot = readRDS("plot_finance_100.rds" )
fin.summary.1 = finance.plot %>% group_by(threshold, method) %>% 
        summarise_each(funs(mean_profit_ratio = round(mean(.), 1), 
                            sd_prof_ratio = round(sd(.), 1)),
                       profit.ratio)

@@

# test each sample against "average" because it's the best
# n.samples function is defined in 'draft.rmd'
index = c("orig.boost","orig.log", "orig.rf","simp.log")
ns = NULL
for(i in seq_along(index)){
        temp = n.samples(data = list(power.fin %>% filter(method == index[i]) %>% .$profit.ratio, power.fin %>% filter(method == 'average') %>% .$profit.ratio), type.calc = 'two.sample', alt = 'two.sided')
        ns[i] = temp
}
ns

# orig.boost needs 6, orig.log needs 4, orig.rf needs 6 and simp.log needs 64 : ) do 100!

@@

Using this logic, if the threshold was 0, all projects were rejected and total profit would equal 0. If the threshold was 1, all projects were accepted and the total profits would be the same as historical figures. Probability results from the two blended models as well as three individual models were tested and because the models require a training and testing set, the process was repeated 100 times with 20 x 5 fold cross validation models to achieve a statistical power of 0.8 (calculated after an initial run of 10 models). The plot below illustrates the results of each model at each threshold with a local polynomial regression line of best fit (loess) and a 95% confidence interval around the mean.

@@

finance.plot = readRDS("plot_finance.rds" )

ggplot(finance.plot, aes(x= threshold, y= profit.ratio, colour = method)) + 
        geom_point(alpha = 0.1, shape = 16) + 
        stat_smooth() +
        scale_y_continuous(breaks = seq(0,110, by = 10)) + 
        scale_x_continuous(breaks = seq(0,1, by = .1)) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2) + 
        geom_hline(yintercept = 0, colour = "darkgray", linetype = 1) + 
        theme(legend.position = "bottom") + 
        scale_colour_discrete(labels = c("average" = "Blend - Average",
                                       "orig.boost"= "Boosted Trees",
                                       "orig.log" = "Logistic Regression",
                                       "orig.rf" = "Random Forest",
                                       "simp.log" = "Blend - Log Reg"))+
        guides(col = guide_legend(nrow = 2)) +
        labs(x = "Probability Threshold", 
             y = "Ratio of Model Profit to Historical Profit (%)",
             title = "Profit Curve for Blended and Individual Models")

@@

The two blended models outperformed the individual models as shown in this table describing each models' optimal thresholds:

@@

fin.summary = finance.plot %>% group_by(threshold, method) %>% 
        summarise_each(funs(mean_profit_ratio = round(mean(.), 1), 
                            sd_prof_ratio = round(sd(.), 1)),
                       profit.ratio)

#for each method find max mean

best.fin = fin.summary %>% group_by(method) %>% 
        filter(mean_profit_ratio == max(mean_profit_ratio)) %>% 
        arrange(desc(mean_profit_ratio))
kable(best.fin)

best.mean = best.fin %>% filter(method == "average") %>% .$mean_profit_ratio
best.sd = best.fin %>% filter(method == "average") %>% .$sd_prof_ratio
best.t = best.fin %>% filter(method == "average") %>% .$threshold

@@

The average of the three individual models performed best with the highest mean profit ratio at `r best.mean`% and the narrowed standard deviation at `r best.sd`%. This means that for the blended average model, if all jobs above the probability threshold 0.55 were rejected, the profits would increase on average by `r best.mean - 100`% in comparison to historical profit data.

For clarity, the plot for the average blended model is shown below:

```{r av_profit_curve, fig.height=10}
ggplot(finance.plot %>% filter(method == "average"), 
       aes(x= threshold, y= profit.ratio, colour = method)) + 
        geom_point(alpha = 0.1, shape = 16) + 
        stat_smooth() +
        scale_y_continuous(breaks = seq(0,110, by = 10)) + 
        scale_x_continuous(breaks = seq(0,1, by = .1)) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2) + 
        geom_hline(yintercept = 0, colour = "darkgray", linetype = 1) + 
        theme(legend.position = "none") +
        labs(x = "Probability Threshold", 
             y = "Ratio of Model Profit to Historical Profit (%)",
             title = "Profit Curve for Blended Average Model")

@@

# Discussion - 10 to 20 pages

>>>>

![This method was implemented by the British government for a rail project which is still under completion](images/lit_review/UKrail.JPG)

@@

The following chapter, Chapter 4, describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting return per dollar, followed by a simpler binary classification problem predicting profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision making scenarios and further applications were considered.

The stages of research outlined in the method were divided into four stages, each of which was assigned a chapter containing the results and discussion of that portion:

* Chapter 5: Variable Selection
* Chapter 6: Regression Models
* Chapter 7: Binary Classification Models
* Chapter 8: Extended Research including model blending, overall profit analysis, research limitations and future work

The results section in each chapter presents key charts and tables that progressively answer the research question: can a statistical or machine learning model based on a company's internal historical project data predict the profitability of a new project and have a positive impact on the bottom line of the business? Then, a discussion finalises each chapter, where the results are examined deeply and analysed in a broader context. The nature of the results are also debated in terms of how surprising they were, their impact on the research hypothesis and limitations as well as suggested future work. The final chapter, the Conclusion, summarises the findings of the project, how the research answered the hypotheses and how the work could be viably applied in industry.

@@

## Method Conclusion

<!-- In summary, to determine whether a  -->
<!-- * obtained data: 4169 jobs from a single case study company over 12 years. Data was de-dentified by an employee of the company -->
<!-- * the data was all from the company CRM. There were three chief sources: -->
<!--     * invoicing data -->
<!--     * timesheet data -->
<!--     * project summary data -->
<!-- * the data was cleaned from mistakes which were generally discovered as outliers  -->
<!-- * many new variables were engineered from the data, often by summarising time series data into a value that is relevant to a final data set of a single job per row -->
<!-- * variable selection: dozens of variables available which were condensed to the most important 10-12. variable importance methods were ANOVA, random forest, and cforests. the overlapping important variables were chosen first as well as some others -->
<!-- * 5 models were tested: baseiline: log reg, naiive bayes. -->
<!-- * sophisticated ML: boosted reg, random frest, bayesian networks -->
<!-- * best were comapred to a model averaged version -->
<!-- * Analysis of impact on bottom line for business if model is integrated into decision making -->
<!-- * link: results -->

# Chapter _ Variable Selection

@@

## Linear Regression

@@

## Random Forests

@@

## Cforest

@@

Once several methods of variable importance analysis had been completed, the separate results were rationalised and final variables were selected for the predictive models.

## Variable Selection Results Summary

Based on the results of the three methods for variable importance, 11 explanatory variables were selected for predictive modelling. They are listed in the table below along with the methods that highlighted each variable as important (randomForest = RF, cForest = CF):

@@

## Variable Selection Discussion

The variable importance results from the three types of models: ANOVA regression, randomForest, and cForest showed more similarities than differences. All three models ranked five variables highly:

* percent of hours by a professional
* project timespan
* number of employees in project team
* total amount invoiced
* business category of client

This was not necessarily expected, particularly between ANOVA and the ensemble tree methods as ANOVA measures linear relationships in constrast to non-linear ensemble trees. Random forests favour variables that are numeric or have many categories and this held true for the analysis. The 6 highest-ranked variables were numeric with one categorical variable, Business category of client, which had 28 categories. The two lowest ranking variables had only 4 categories. It was expected that cForests would rank the categorical variables with only 4 categories more fairly. This was the case with Discipline and it recieved a mid-level ranking. Business category of the client and Broad business type however, were ranked similarly to randomForests.

The qualities of the higher ranked variables were not surprising, and were generally centered around who did how much of the work internally, general client characteristics, project timespan, and project category. It was expected that Billing Type would play a more important role in predicting profitability as it describes the structure of the incoming fees. For this reason, and with advice from the case study business, the Billing Type was kept in the model.

For construction projects and IT software projects, the literature indicated size was the most important variable predicting effort. Size could mean size of a building or function point (level of functionality of the software) [@Shepperd1996][@Finnie1997][@Pai2013]. In this case study, profitability is the dependent variable, not total effort and the size variable plays a different role. Instead of contributing to predicting the total effort, it is describing whether the size of a project correlates with how profitable the project is. The closest estimate of size in this case study is the invoiced amount variable which did rank as very important. The exact invoiced amount is difficult to guess before a project begins, and is part of the problem this study is addressing. Therefore, the invoiced amount was binned into categories that a manager could much more confidently choose as an estimate. The other variables indicated as important in this study  (including the number of employees on a project, client business type and percent of hours by certain types of employees) have not been strongly represented in the literature to date. This is because the variables were easily calculated from internal CRM data but would not be easily obtained externally from projects across many companies or regions.

The 11 explanatory variables from table __ were progressed to the next stage where they were used to build predictive models. The outcomes of the regression and categorical models are summarised and discussed in the following two chapters.

# Chapter _ Regression Models

Prediction of 'return per dollar' as a regression problem was attempted first followed by prediction of profit or loss as a binary classification problem. Linear Regression, Random Forests and Boosted Tree methods were applied to both problems with varying success.

## ANOVA

@@

were statistically significant. Two types of ANOVA models were run. The first used only a 'core' list of 6 variables that were largely complete and any relevant interaction terms.

* Total amount invoiced 
* Discipline 
* Timespan
* Team size
* % of hours completed by a professional level employee
* Client business category

The second method was developed to overcome the problem of missing data. When all variables were combined, only 15% of the project cases had complete data and therefore the ANOVA test would eliminate 85% of the data upon execution. Imputing the missing data was not desirable at that stage due to the volume of missing categorical data and an inability to ensure the imputed data was not hampering results. The second method was a more complicated procedure that required dozens of models but allowed variables with missing data to be included. First, a project was randomly selected (a row in the dataset) which would have certain variables complete. The data was then filtered for all projects that had at least the same columns complete, and an ANOVA model was built on this reduced data set. This method allowed variables with a larger proportion of missing data to be included while not attempting to analyse all variables at once.

@@

results.A2 = readRDS("report_data/reg_lm.rds")
results.A2 <- dplyr::filter(results.A2, diff>-15)
# n.samples(data = results.A2$diff)

@@

## Random Forest

```{r load_rf_results}
results.rf.core = readRDS("report_data/reg_rf_core.rds")
# n.samples(data = results.rf.core$diff)

results.rf2 = readRDS("report_data/reg_rf.rds")
# n.samples(data = results.rf2$diff)

@@

The two regression models predicting 'return per dollar' resulted in high mean RMSE's ranging from \$`r mean(results.rf.core$final.RMSE) %>% round(2)` to \$`r mean(results.A2$final.RMSE) %>% round(2)`. The next section discusses possible reasons for the poor results and the decision to halt progression of regression models.

## Regression Models Discussion

Both ANOVA and randomForest models produced a similar lowest RMSE value, but performed differently when directly comparing models built on the same list of variables. Both methods predicted at best within an average of \$0.48 for the response variable, 'return per dollar', using either core variables or a sample of core and less complete variables. Random Forest predictictions were much more accurate than ANOVA's with a mix of core (almost complete) and less variables while randomForest and ANOVA performed similarly with the core variables only. There was much less data available for training the models that used less-complete variables so it was interesting that randomForest performed considerably better (RMSE of \$0.50 for randomForest vs. \$2 for ANOVA). The main difference between ensemble tree methods and ANOVA is that ANOVA assumes linear relationships between variables (some of which are normalised) whereas trees are capable of capturing non-linear relationships. It is possible the less complete variables had non-linear relationships with the response variable. RandomForest is also an ensemble method while ANOVA is not. Each randomForest is the average of 500 single trees, whereas each ANOVA model is a single model only. 50 randomForest models produce 500 times more models than 50 ANOVA models so the averaging effect may have eliminated outliers for better overall performance.

If the best RMSE achievable at that point was \$0.48, what does that imply in terms of the business decision? The magnitude of the RMSE can be evaluated from the end-user's perspective (the manager who is proposing fees for new projects). To give a feel for typical project performance, refer the plot below of historical 'return per dollar' values:

The 95% confidence interval for this data is \$`r (mean(all8a$return.pdol)- qnorm(0.975)*sd(all8a$return.pdol)) %>% round(2)` to \$`r (mean(all8a$return.pdol) + qnorm(0.975)*sd(all8a$return.pdol)) %>% round(2)`. If a project returns \$0 for every dollar spent (return per dollar = \$0), the project has earned only enough to cover the cost of doing the project (salary costs and business costs). If the return per dollar is -\$0.50, the project has earned enough to cover half of the costs of doing the job. Finally, if the return per dollar is \$1, the project earned double what it cost to perform. Using this reasoning, if model predictions have an average error of \$.50 for return per dollar, this represents an error in revenue of half what the project cost to complete. Unfortunately, this is clearly not informative enough to bring into the business decision-making process.

Further regression methods could have been persued, however randomForest is a high-performing algorithm and was far from providing usable results. It is likely that either not enough data was available or that the given variables did not capably explain all the factors affecting return per dollar. Each row of data is a project, which can take the company from a few weeks to sometimes years to complete. Therefore, more data cannot be easily accessed or created. It was decided categorical predictions should be pursued next as predicting whether a job will be profitable or not is a simpler task and could more effectively leverage the limited data.

This conclusion differs to case studies in the literature which tend to predict total cost/effort as a regression problem [@Kim2004][@Chan2005][@Akintoye2000]. This is likely because the literature dealt with substantial projects which are rarer and impact governments and large companies. A simple profit or loss prediction would not be useful. The case study company on the other hand may find a profit/loss binary prediction useful because they complete hundreds of small projects a year. If binary classification is successful, applications of the predictions must be explored along with contributions to the company decisions and profits.

In summary, because of the poor results when predicting return per dollar, further development of regression models was deferred in favour of binary classification models. Binary classification is a simpler task than regression because it requires less precision. The challenge is how to gainfully apply the results to the business scenario.

# Chapter _ Binary Classification Models

The cost estimation problem analysed in the literature review had been converted to a profitability estimation problem and was finally converted to a binary profit/loss prediction. The predictive formula was re-structured so that the 11 explanatory variables predicted a new response variable: profitable project (or not). The original 'return per dollar' response variable was transformed by converting all 'return per dollar' values less than or equal to zero to 'loss making' jobs and 'return per dollar' values greater than 1 to 'profitable' jobs. The following section summarises results from the trialled methods.

## Results from Five Methods

The five prediction methods, Boosted Trees, Naive Bayes, Logistic Regression, Random Forests and Bayesian Networks, were intially developed without imputing data, using the maximum amount of data possible depending on the method. Boosted Trees and Naiive Bayes are able to process data with missing values, so all data could be input [@gbm; @e1071]. Logistic Regression, Random Forest and Bayesian Networks require datasets without missing values so a subset of complete data was used. It was expected that Boosted Trees would perform best as they are known to have powerful predictive capabilities and can use incomplete data. For this reason, 4 x 5-fold runs of each method were performed (20 models each) and statistically compared against Boosted Trees' results. A power calculation was then used to calculate the number of runs required to achieve a statistical power of 0.8.

@@

Before the plot could be made, the Boosted Tree algorithm parameters were tuned to optimise the AUC outcomes. The caret package was used to tune the following parameters [@caret]:

@@

<!-- image below generated from file rf_log_naiive_auc.rmd -->
<!-- lines 683 to 689 -->

@@

reduc.totAUC <- readRDS("report_data/reduc_df_auc.rds")

@@

## Binary Classification Discussion

Testing Boosted Trees, Random Forests, Naive Bayes, Logistic Regression and Bayesian Networks on the binary classification problem showed that Boosted Trees, Logistic Regression and Random Forests performed best (according to AUC). Bayesian Networks are not particularly popular and require discretised variables so it was expected that this algorithm would not perform as well as highly regarded methods such as Boosted Trees and Random Forests. Naive Bayes and Logistic Regression were included as baseline models and it was expected that the more complex models would outperform these. Therefore it was surprising that results from 100 Logistic Regression models were not statistically lower than Boosted Trees.

A possible explanation for this, according to the literature, is that there was not enough data for the ensemble trees to learn the complex decision rules at which they excel. Trees tend to overfit the patterns in a smaller training set. Logistic regression on the other hand is capable of only one decision boundary (which does not have to be parallel to the variable axes) and is not prone to overfitting [@Perlich2003]. This may explain Logistic Regression's comparitively high performance on the case study's smaller dataset with many variables.

Boosted trees, Random Forests and Logistic Regression all had mean AUC values around 0.775. The tests were repeated using imputed data and again the means sat around 0.775. The variables were not particularly well suited for imputing so this result is logical and could have even diminished the mean AUC. For example, many of the imputed variables were categorical such as job category, percent of job completed by a professional level employee, and the position of the majority worker on a job. The benefit of having an imputed dataset that performs acceptably well is that each model is fed the full data set and therefore produces predictions for the full data. Beforehand, randomForest and Logistic Regression models were limited to a complete set of original data which implied a reduced data set. With complete model predictions, the high performing models could be blended to test whether better predictions were possible. For this reason, the imputed data models were propogated forward to the development of blended models.

In summary, the individual binary classification models performed well above random chance (AUC = 0.5) and it can be expected that implementing a model would improve a manager's ability to predict whether a new job will be profitable or loss-making. Whether the model is worth implementing in the work place is dependent on the extent to which the algorithm would improve 'bottom line' profits for the business and if the model actually affects decisions in practise. Results from model blending and overall profit impact are presented and discussed in the next chapter.

# Chapter _ Extended Analysis

Blended models combine the predictions from each higher-performing individual model to create averaged or 'blended' predictions. In this case, the randomForest, Boosted Trees, and Logistic Regression models were chosen and as outlined in the method, 6 techniques for model blending were trialled ranging from simple to complex. The simplest way averaged the probability output (a number between 0 and 1) of all three models. The next two simplest methods consisted of building a Logistic Regression and Boosted Trees model from the three individual model outputs only. Feature Weighted Linear Stacking (FWLS), randomForest and Boosted Trees were also tested as blending methods in a more complex scenario where the predictions from each model became additional variables to the original explanatory variables (meta-features). The model predictions were interacted with each original variable so that models that performed more strongly under certain meta-feature conditions could be weighted as such. Some meta-features may have been dispensable in the blending process, so in order to maintain simple comprehensible models, a variable importance study was conducted again.

### Variable Selection: Blended Model

@@

An ANOVA regression found the F-statistics from the following variables were significant in rejecting the null hypothesis for p-value < 0.10:

@@

* Timespan * results of logistic regression model
* Total amount invoiced * results of logistic regression model
* Total amount invoiced * results of randomForest model
* Client Business type * results of boosted tree model

@@

* Total amount invoiced for the job
* Timespan
* Team size
* Client business category
* Project category
* Results from each individual model

@@

Random Forests and Boosted Trees passively interact explanatory variables, but for logistic regression, these must be manually input. Only interaction terms with statistically significant F-statistics from the ANOVA test were included in order to reduce noise.

### Comparison of Blended Models

@@

auc.list <- readRDS("report_data/blend_auc2.rds")
means = lapply(auc.list, mean) %>% unlist
greater = means[means> means["orig.log"]] %>% names
auc.melt = melt(auc.list)
auc.melt$greater = ifelse(auc.melt$L1 %in% greater, TRUE, FALSE)

@@

```{r, fig.width = 8, fig.height=5}

@@

## Profit Analysis

The first research question, whether machine learning and statistical techniques can predict project profitability based on historic data, has been answered. The measure of profitability could not be modelled but simplifying the profitability prediction to a binary (profit/loss) model was successful. To answer the second research question, whether the model would improve the business' profitability, a profit analysis was performed.

The models output a 'probability' between 0 and 1 that each project will be a loss-making job (probability = 1 indicates a loss-making job). This leaves the question, at what probability would a decision-maker round the probability up to 1 or down to 0? And what business decision would then be made? To find the threshold point for rounding, an experimental business-scenario was tested. At a given threshold, all projects calculated above the threshold were considered too risky, and were rejected. All profits and losses from these projects were forfeited, while the profits from the remaining jobs (below the threshold) were summed to give a revised total profit.

@@

# How to achieve statistical power of 0.8?
finance.plot = readRDS("report_data/plot_finance.rds" )
fin.summary.1 = finance.plot %>% group_by(threshold, method) %>% 
        summarise_each(funs(mean_profit_ratio = round(mean(.), 1), 
                            sd_prof_ratio = round(sd(.), 1)),
                       profit.ratio)

@@

# test each sample against "average" because it's the best
# n.samples function is defined in 'draft.rmd'
index = c("orig.boost","orig.log", "orig.rf","simp.log")
ns = NULL
for(i in seq_along(index)){
        temp = n.samples(data = list(power.fin %>% filter(method == index[i]) %>% .$profit.ratio, power.fin %>% filter(method == 'average') %>% .$profit.ratio), type.calc = 'two.sample', alt = 'two.sided')
        ns[i] = temp
}
# ns

# orig.boost needs 6, orig.log needs 4, orig.rf needs 9 and simp.log needs 64 : ) do 100!

@@

Using this logic, if the threshold was 0, all projects were rejected and the total profit would equal \$0. If the threshold was 1, all projects were accepted and the total profits would be the same as historical figures. This profit calculation was made on a range of thresholds between 0 and 1 at 0.05 increments. The total profits were plotted for each threshold and joined to make a profit curve where at an optimal threshold, the total profits were highest (refer plot below).

Profit curves for the three simple blended models as well as the three individual models were calculated. However, the total profit coordinates were based on probability outputs in test datasets. This means that the profit curve for the simple logistic regression blend (for example) would be different for different training/testing data divisions. I.e. the total profits at each point in the profit curve depends on the model parameters that were calculated using the training data and which projects were analysed in the test data. Therefore, the process of training a blended model, then calculating the profit curve from the test data was repeated multiple times to create a distribution of profit curves for each method (blended and individual).

To start, 10 curves for each method were calculated (2 x 5-fold cross validation). Two-sampled power calculations (for a power of 0.8) were performed between the results of each method at their optimal threshold against the best performing method (simple average blend). For example, the optimal threshold for the simple average blended model was 0.55 while the optimal threshold for the original boosted tree model was 0.60. The 10 'total profit' values at these two thresholds were compared in the two-sample power calculation to determine the number of samples required to achieve a power of 0.8. The maximum required number of samples was 64 so 100 models were of each method were built (20 x 5 fold cross validation).

The plot below illustrates the results of each model at each threshold with a line joining the mean value at each threshold for each method. The grey ribbon illustrates the 95% confidence interval around the mean at each threshold for the 100 models.

@@

finance.plot = readRDS("report_data/plot_finance_100_3.rds" )
finance.plot$method <- factor(finance.plot$method, 
                                    levels=c("average", "simp.log","simp.boost",
                                             "comp.boost","comp.rf", "FWLS","orig.log",
                                             "orig.rf","orig.boost"))

ggplot(finance.plot, aes(x= threshold, y= profit.ratio, colour = method)) + 
        # geom_point(alpha = 0.005, shape = 16, size = 1) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2) + 
        geom_hline(yintercept = 0, colour = "darkgray", linetype = 1) +
        stat_summary(fun.y = mean, 
                     fun.ymin = function(x) mean(x) - 1.96*sd(x),
                     fun.ymax = function(x) mean(x) + 1.96*sd(x),
                     fill = "gray", geom= "ribbon", alpha = 0.7, colour = NA) +
        stat_summary(fun.y = mean, geom="line") +
        scale_y_continuous(breaks = seq(0,110, by = 10)) + 
        scale_x_continuous(breaks = seq(0,1, by = .1)) +  
        theme(legend.position = "none") +
        labs(x = "Probability Threshold", 
             y = "Ratio of Model Profit to Historical Profit (%)",
             title = "Profit Curve for Blended and Individual Models") + 
        facet_wrap(~method, 
                   labeller = labeller( method = c("average" = "Simple Average Blend",
                                       "orig.boost"= "Orig. Boosted Trees",
                                       "orig.log" = "Orig. Logistic Regression",
                                       "orig.rf" = "Orig. Random Forest",
                                       "simp.log" = "Simple Log. Reg. Blend",
                                       "simp.boost" = "Simple Boosted Tree Blend",
                                       "comp.boost"= "Complex Boosted Tree Blend",
                                       "comp.rf"= "Complex randomForest Blend",
                                       "FWLS" = "FWLS"
                                       )))

@@

The two blended models outperformed the individual models as shown in the plot above with higher profit ratios as well as tighter confidence intervals. The table below denotes each models' optimal thresholds and the corresponding increase in total profits:

@@

fin.summary = finance.plot %>% group_by(threshold, method) %>% 
        summarise_each(funs(mean_profit_ratio = round(mean(.), 1), 
                            sd_prof_ratio = round(sd(.), 2)),
                       profit.ratio)
#for each method find max mean

best.fin = fin.summary %>% group_by(method) %>% 
        filter(mean_profit_ratio == max(mean_profit_ratio)) %>% 
        arrange(desc(mean_profit_ratio), sd_prof_ratio) %>% 
        slice(1) %>% ungroup() %>% 
        arrange(desc(mean_profit_ratio), sd_prof_ratio)

best.fin <- best.fin %>% mutate(lower_bound_95CI = (mean_profit_ratio - 1.96*sd_prof_ratio)%>% round(1))
best.fin <- best.fin %>% mutate(upper_bound_95CI = (mean_profit_ratio + 1.96*sd_prof_ratio)%>% round(1))

kable(best.fin)
# saveRDS(best.fin, "report_data/best_fin.rds")

best.mean = best.fin %>% filter(method == "simp.log") %>% .$mean_profit_ratio
best.sd = best.fin %>% filter(method == "simp.log") %>% .$sd_prof_ratio
best.t = best.fin %>% filter(method == "simp.log") %>% .$threshold

@@

The simple logistic regression blend performed best with the highest mean profit ratio at `r best.mean`% with a standard deviation of `r best.sd`%. This means that for the simple blended logistic regression model, if all jobs above the probability threshold `r best.fin %>% filter(method == "simp.log") %>% .$threshold` were rejected, the profits would increase on average by `r best.mean - 100`% in comparison to historical profit data.

For clarity, the plot for the simple blended logistic regression model is shown below:

```{r av_profit_curve}
ggplot(finance.plot %>% filter(method == "simp.log"), 
       aes(x= threshold, y= profit.ratio, colour = method)) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2) + 
        geom_hline(yintercept = 0, colour = "darkgray", linetype = 1) + 
        geom_point(alpha = 0.1, shape = 16) +
        stat_summary(fun.y = mean, 
                     fun.ymin = function(x) mean(x) - 1.96*sd(x),
                     fun.ymax = function(x) mean(x) + 1.96*sd(x),
                     fill = "gray", geom= "ribbon", alpha = 0.7, colour = NA) +
        stat_summary(fun.y = mean, geom="line") +
        scale_y_continuous(breaks = seq(0,110, by = 10)) + 
        scale_x_continuous(breaks = seq(0,1, by = .1)) + 
        theme(legend.position = "none") +
        labs(x = "Probability Threshold", 
             y = "Ratio of Model Profit to Historical Profit (%)",
             title = "Profit Curve for Simple Logistic Regression Blended Model")

@@

## Extended Analysis Discussion
